Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-21 paper.qmd
Update paper.qmd
Revise Nix package manager section for clarity and detail
Enhanced the section on the Nix package manager to provide a more detailed explanation of its features, advantages, and cross-platform support. Improved clarity on dependency management, reproducibility, and the challenges of using Nix.

[Nix]{.pkg} is a package manager designed to install and build software in a
fully reproducible manner. As of this writing, it provides access to over
120000 packages, including nearly all of CRAN and Bioconductor. This allows
users to install not only [R]{.proglang} itself but also all packages required
for a given project. Although [Nix]{.pkg} is the package manager of the NixOS
Linux distribution, it can also be installed as a standalone tool on other Linux
distributions, macOS, and the Windows Subsystem for Linux^[Effectively treating
non-NixOS Linux distributions and Windows as equivalent platforms in practice.].

The advantage of using [Nix]{.pkg} to install [R]{.proglang} packages, instead
of the standard `install.packages()` function, is that [Nix]{.pkg} ensures all
dependencies of each package are installed, whether they are other
[R]{.proglang} packages or system-level libraries.

For example, the [xlsx]{.pkg} package requires [Java]{.proglang} to be
installed. On some systems, installing [Java]{.proglang} may be difficult or
even impossible manually. With [Nix]{.pkg}, however, the user only needs to
declare that [xlsx]{.pkg} is required for the project; [Nix]{.pkg} automatically
installs and configures [Java]{.proglang} as a dependency. This works because
the maintainers of the [R]{.proglang} language packages for [Nix]{.pkg} have
declared [Java]{.proglang} as a dependency of [xlsx]{.pkg}, allowing the process
to happen seamlessly from the end-user perspective.

[xlsx]{.pkg} can thus be referred to as a component closure, and quoting
@dolstra2004nix:

@@

But how does [Nix]{.pkg} achieve this reproducibility? Where do these packages,
or *closures*, come from? When installing a package with [Nix]{.pkg}, an
expression written in the [Nix]{.pkg} language is downloaded from the `nixpkgs`
GitHub repository and evaluated. These expressions define *derivations*, which
describe how to build a package: specifying its dependencies, the commands to
build and install it, and the resulting output. Typically, a derivation
downloads the source code, compiles it, and outputs a binary. Derivations are
highly flexible; by writing custom derivations, users can define and build
reproducible environments instead of a single program. The dependencies of each
derivation are themselves defined in other expressions and are automatically
installed if needed. In the case of [R]{.proglang}, maintainers declare the
appropriate dependencies to create fully self-contained component closures.

Why is installing software with [Nix]{.pkg} reproducible? Because the entire set
of [Nix]{.pkg} expressions is hosted on GitHub, it is possible to reference a
specific commit of `nixpkgs`, a process called *pinning a revision*. Pinning a
revision ensures that all packages installed by [Nix]{.pkg} are always the exact
same versions, regardless of when in the future they are built, since the
expressions downloaded correspond to that specific commit of `nixpkgs`.

Pinning is crucial, but it is not the only reason [Nix]{.pkg} supports
reproducibility. [Nix]{.pkg} is a functional package manager: it applies
principles from functional programming, disallowing side effects and global
variables, and ensuring that the output of a build is always the same given the
same inputs, regardless of time or location. A side note: while this functional
approach greatly enhances reproducibility, it can make writing derivations more
complex, particularly for packages that need to download assets during
installation (e.g., some Bioconductor packages like [musData]{.pkg}). However,
this concern is primarily for package maintainers, not end-users.

Additionally, [Nix]{.pkg} supports multiple versions, or *variants*, of a
package on the same system. Each variant has a unique identifier, allowing
multiple versions of [R]{.proglang} to coexist and ensuring the correct version
is used for each project. For a more technical discussion of [Nix]{.pkg}, we
refer to @dolstra2004nix.

With [Nix]{.pkg}, it is possible to effectively replace both [renv]{.pkg} and
[Docker]{.pkg} for [R]{.proglang} projects—or, in the case of
[Python]{.proglang}, to replace `requirements.txt` files and virtual
environments. [Nix]{.pkg} also supports building multi-language environments,
which can include [R]{.proglang} and [Python]{.proglang}, a \LaTeX
distribution or any other of the 120,000 available tools. This enables the
creation of a complete, project-specific, and reproducible environment that can
be used interactively or non-interactively. As long as the `nixpkgs` GitHub
repository remains online, the environment can be rebuilt in the future to rerun
analyses reliably.

However, [Nix]{.pkg} has a steep learning curve. It is a complex system that
comes with its own programming language, also called [Nix]{.proglang}, designed
to solve the challenging problem of declaratively defining how software is built
and configured. This ensures that installations are fully reproducible across
operating systems and hardware platforms. To make [Nix]{.pkg} more accessible to
[R]{.proglang} users that wish to use project-specific and reproducible
development environments, we developed the [rix]{.pkg} package.

>>>>

[Nix]{.pkg} is a powerful, cross-platform package manager designed for installing 
and building software in a fully reproducible and reliable manner. Unlike traditional 
package managers, [Nix]{.pkg} emphasizes immutable infrastructure and a functional
approach, which significantly enhances the consistency of computational environments.
As of writing, the primary Nix package collection, `nixpkgs`, provides access to over 
120,000 packages, including nearly all of CRAN and Bioconductor. 
This extensive coverage allows users to install not only [R]{.proglang} itself but 
also all its required packages and system-level dependencies for any given project.

While [Nix]{.pkg} is the default package manager for the NixOS 
Linux distribution, it can be installed as a standalone tool on other Linux 
distributions, macOS, and effectively on Windows via the Windows Subsystem for 
Linux (WSL2), treating these as equivalent platforms in practice.

A key advantage of using [Nix]{.pkg} to install [R]{.proglang} packages, as 
opposed to the standard `install.packages()` function, lies in its comprehensive 
dependency management. [Nix]{.pkg} ensures that all dependencies of each package are
installed, irrespective of whether they are other [R]{.proglang} packages or 
underlying system libraries.

Consider, for example, the [sf]{.pkg} package, which is fundamental for spatial data analysis in [R]{.proglang}.
[sf]{.pkg} relies on several complex external system libraries such as [GDAL]{.proglang}, 
[GEOS]{.proglang}, and [PROJ]{.proglang}. Manually installing and configuring these libraries 
to be compatible with [sf]{.pkg} can be challenging and platform-specific. With [Nix]{.pkg}, 
however, the user merely needs to declare [sf]{.pkg} as a project requirement. 
[Nix]{.pkg} then automatically installs and configures all necessary system libraries
as transitive dependencies. This seamless process is possible because the maintainers of the 
[R]{.proglang} programming language for [Nix]{.pkg}
have declaratively specified these dependencies for [sf]{.pkg} in `nixpkgs`, allowing the entire software 
stack to be provisioned effortlessly from an end-user perspective.

This concept is central to [Nix]{.pkg}'s design, where packages are referred to as *component closures*. 
As @dolstra2004nix explains:

@@

The core of [Nix]{.pkg}'s reproducibility lies in its functional package management paradigm
and its unique approach to builds. When installing software with [Nix]{.pkg}, it evaluates
an *expression* written in the [Nix]{.proglang} language. These expressions define *derivations*,
which are declarative blueprints describing how to build a package. A derivation specifies all inputs: 
source code, build commands, and crucially, all direct and transitive dependencies. This declarative 
nature ensures that every build process is fully specified and independent of the environment where 
it's executed. Furthermore, all [Nix]{.pkg} expressions for official packages are hosted in the `nixpkgs`
GitHub repository. By referencing a specific commit of `nixpkgs` (a process known as *pinning a revision*) users 
guarantee that all packages installed are derived from the exact same set of build instructions.
This means that if an environment is built today using a pinned revision, it will produce the 
identical set of software versions and configurations if rebuilt tomorrow, next year, or on a 
different machine, as long as the `nixpkgs` repository remains accessible.

Pinning is crucial, but it is not the only reason [Nix]{.pkg} supports reproducibility.
[Nix]{.pkg} is a *functional* package manager in the programming sense: it treats software builds as pure functions.
Given the same inputs (source code, dependencies, build scripts), a [Nix]{.pkg} build will always produce the 
identical output, regardless of when or where it's executed. This is achieved by disallowing side effects and 
ensuring that builds operate within hermetic (isolated) environments, with no hidden dependencies on global system 
state. While this functional approach greatly enhances reproducibility, it can sometimes introduce complexity for
package maintainers, especially for software that needs to download external assets during installation (e.g., 
certain Bioconductor packages). For end-users, however, this complexity is largely abstracted away. Additionally, 
[Nix]{.pkg} supports multiple versions or *variants* of a package on the same system. Each package installed into
the "Nix store" (typically in `/nix/store`) has a unique identifier derived from a hash of all its inputs. This 
isolation ensures that projects always use their intended software versions, allowing multiple versions of
[R]{.proglang} to coexist and preventing "dependency hell" and global environment pollution.

For a more in-depth technical discussion of [Nix]{.pkg}'s design principles, we refer readers to @dolstra2004nix.

By leveraging these principles, [Nix]{.pkg} can effectively replace and integrate the functionalities of 
tools like [renv]{.pkg} and [Docker]{.pkg} for [R]{.proglang} projects, or `requirements.txt` files 
and virtual environments for [Python]{.proglang} projects. Furthermore, [Nix]{.pkg} excels at building
multi-language environments, which can seamlessly combine [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang},
a [LaTeX]{.proglang} distribution, or any of the numerous other tools available in `nixpkgs`. 
This enables the creation of a truly complete, project-specific, and deeply reproducible environment
that can be used interactively for development or non-interactively for automated analysis. 
As long as the `nixpkgs` repository (or a suitable fork, as discussed in Section @sec-fork) remains accessible,
the environment can be rebuilt reliably in the future.

[Nix]{.pkg} provides strong cross-platform support across major operating systems. It runs natively on Linux 
(x86_64) with full functionality and optimal integration. On macOS (Intel and Apple Silicon), however, 
reproducibility is less reliable due to dependencies on impure system frameworks and Xcode toolchains outside
the Nix store. As a result, macOS users may need to update project pins after system or Xcode upgrades to 
restore builds. On Windows, [Nix]{.pkg} works well through the Windows Subsystem for Linux (WSL2), though graphical
applications may need extra setup. Nix environments on WSL can be used interactively via VS Code or Positron 
for a smooth development experience.

While [Nix]{.pkg} is a powerful system, it comes with notable challenges. The steepest is its learning curve:
[Nix]{.pkg} uses its own declarative language (also called [Nix]{.proglang}), which can be difficult for 
those new to functional or declarative paradigms. Complexity becomes most apparent when writing custom 
derivations or troubleshooting intricate build issues. Build times can also be long, especially for large
environments, when pre-built binaries aren’t available. Additionally, because [Nix]{.pkg} stores all package
outputs immutably in the Nix store, disk usage tends to be higher than with traditional package managers. 
Finally, while `nixpkgs` is extensive, some new or niche packages may be missing or fail to build across platforms.

To make these capabilities more approachable for [R]{.proglang} users seeking reproducible, 
project-specific environments without learning the full [Nix]{.proglang} language, I created the 
[rix]{.pkg} and [rixpress]{.pkg} packages.
